Leadership across cultures and contexts: an exploratory analysis

### 2020-05-11 leadership_across_cultures_contexts.Rmd
Working on discussion section

Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported, suggesting leadership to be a high risk, high return strategy (see Figure \@ref(fig:varsupportcb)). The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to receive mating benefits, relative to leaders or other types benefits. Our results cannot determine if the benefits to leaders and followers outweighed their costs, but they do shed light on the social and biological pressures associated with leadership and social hierarchy. For both leaders and followers, individuals stood to lose what they were able to gain. The major difference between leaders and followers lay in leaders' opportunities for increased social status and mating opportunities, relative to followers. See Figure \@ref(fig:varsupportcb).

All evolutionary models of leadership propose that the fitness benefits of leading outweigh the costs [e.g., @van_vugt_leadership_2008; @kantner_identifying_2010]. The relative abundance of evidence for leader and follower benefits vs. costs ($OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$ and $OR=`r signif(follower_bc_ratio$benefit_cost_OR, 2)`$, resp.), while not providing evidence of net fitness benefits, is nevertheless encouraging, especially since leaders and followers both had more evidence of mating and territory benefits than costs, which are key contributors to fitness, and leaders also had more evidence for inclusive fitness, material and social benefits, relative to costs. These results provide a framework for further work on evolutionarily relevent costs and benfits of leadership, and suggests leaders-follower dynamics should continue to be prioritized in evolutionary models of demography, group dynamics, social complexity and inequality, cooperation and punishment, and resource exchange [c.f., @bird_variability_2019; @ember_our_2018; @micheletti_modelling_2020; @smith_cultural_2020; @birch2017philosophy; @marlowe_more_2008; @henrich_markets_2010; @mace_co-evolution_1998; @mace_phylogenetic_2005; @mattison_evolution_2016] See Figure \@ref(fig:benefitcostratio).

@@

Relationships between leadership and high status, including causality and dependence, are puzzling and currently the focus of much work in evolutionary leadership studies [e.g., @van_vugt_dual_2019; @smith_leadership_2020; @cheng_why_2020; @cheng_dominance_2019; @von_rueden_dynamics_2019]. Drawing on a comparative-evolutionary perspective, @van_vugt_dual_2019 emphasize that although leadership and high social status do often covary, leaders are not always high-status individuals, exemplified when low-ranking individuals in nonhuman species lead group movement or when younger and lower status men and women initiate collective actions (as examples the authors point to raids in small-scale warfare and widespread social-political movements). @cheng_dominance_2019 and @cheng_why_2020, emphasize, however, that social hierarchies form principally in response to inter-individual variation in status, which may be gained via distinct pathways. Thus, according to @cheng_why_2020, although all leaders are not necessarily high status individuals, especially across nonhuman species, among humans high social status may most often be a necessary but insufficient condition for achieving and maintaining positions of leadership. In our view, social context is critical to these discussions [see @wiessner_power_2010]. Leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social "statuses." The "low status" male who initiates an inter-group raid may not be widely respected nor a candidate for community leadership, however, he may be high status among his peers and within what @garfield_review_2019 [p. 71] described as the "intragenerational social hierarchy."

Our results offer some insight into relationships between leadership and social status across cultures and contexts. As mentioned, high status is the most common quality of leaders (Figure \@ref(fig:varsupport)) and the second most common benefit of leadership and the loss of status the most frequent cost (Figure \@ref(fig:varsupportcb)). 
<!-- mentioned in this sample of the ethnographic record which includes discussions of chiefs, work group leaders, shamans, and age-sets, from a culturally diverse sample including foraging groups and kingdoms.  -->
This lends support to the notion of high social status often being a necessary condition of leader emergence [@cheng_why_2020]. The dimensions of leadership predictive of high status leaders suggest the phenotype of such leaders is linked to specific competencies, including experience, decision-making, intelligence, and inter-personal skills, as well as older age and conformity to social norms, which does support claims of independence by @smith_leadership_2020 [see @von_rueden_multiple_2008]. High status leaders are able to direct benefits to their close relatives, likely in part to their greater material and social wealth, which implicates leadership dynamics as a driver of the evolution and emergence of inequality [c.f., @sanday_female_1981; @shenk_status_2016; @price_social_2010; @borgerhoff_mulder_pastoralism_2010; @mattison_evolution_2016; @bowles_emergence_2010]. Our results did not strongly implicate high status with particular functions, with the exception that high status leaders are unlikely to enforce sanctions on group members. See Figure \@ref(fig:highstatusplot)B.

Although evidence for high status leaders did not substantially vary by subsistence or group structure, this null result suggests relationships between leadership and social status are not substantially context-dependent or specific to greater social complexity. Speculatively, if anything leadership among lower levels of social organization and less politically complex societies (i.e., kin groups and hunter-gatherers) is more likely to be associated with high social status, relative to other groups (see \@ref(fig:groupstatus)A and \@ref(fig:groupstatus)B). If so, this trend could be due to the institutionalized and overt or ascribed social distinctions among more politically complex populations and the importance of achieve statuses among less politically and socially complex groups. Nonetheless, these findings question the informed speculation by @smith_leadership_2020 that, "leadership may therefore be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." However, our text analytic results, which relied on the semantic context of text records to predict the presence of high status leaders generally identified terms related to status and prestige (i.e., *respect*, *prestige*, *honor*) and possibly dominance or proscribed hierarchy (i.e., *rank*, *position*), as stronger predictors of high status leader than terms associated with leadership per se (i.e., *influence*, *authority*, *headman*), which lends some support to the distinction between status and leadership highlighted by @smith_leadership_2020 (though we caution from drawing strong interpretations from these text-analytic results, see Figure S\@ref(fig:highstatusplot)). Overall, these results suggest associations between leadership and high status are ubiquitous across human societies, including among kin groups and among hunter-gatherers and that while effective leaders across a wide range of social contexts are the recipients of increased social status for their valuable services, ineffective leadership is associated with reduced social status.

Although discussions of female leaders were rare, the presence of female leaders were also associated with high social status (see Figure \@ref(fig:groupstatus)).

@@

Only a small fraction of the ethnographic record sampled here discussed female leaders. Our results do however provide some insight on female leadership and sex-differences in leadership across cultures and contexts (though we emphsize the limitations of our data). Female leaders were more likely to emerge in kin groups and residential subgroups, consistent with theoretical arguments emphasizing women as leaders of families  [e.g., @garfield_evolutionary_2019; @hagen_leadership_2019] and previous cross-cultural studies [e.g., @yanca_female_2004]. Texts describing female leaders exclusively also provided a relatively higher proportion of total leader benefits (Figure \@ref(fig:leadertotalbenefits)) and descriptions of high status leaders (Figure \@ref(fig:groupstatus)) compared to descriptions of leaders that were not sex-specific (holding other measures constant; although this effect was similar to benefits to and high status of male-leader texts). This suggests, much like male leaders who are more often political leaders, female leaders are also high status individuals and receive substantial benefits for their services.

Across cultures and contexts, when ethnographic texts described female leaders exclusively, they were relatively more likely to be described as married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits (Figure S\@ref(fig:heatmapsex)). This suite of characteristics is consistent with discussions of high status first-wives who gain and maintain increased social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985], and supports the view of positive assortative mating of high status leaders @garfield_investigating_2019 documented among elected leaders in a small-scale society with relative gender-egalitarianism. Evidence for male leaders (relative to evidence for female leaders or evidence that was not sex-specific) was indeed linked to group-level and political functions and male leaders were relatively more likely to provide social services, control political structure, and were at greater risk of losing their high social standing. Female leaders, relative to males, were also more likely to lack coercive authority, which is consistent with much of leadership studies suggesting dominant and aggressive women are less likely to be preferred as leaders, unlike dominant men who are sometimes preferred as leaders [@garfield_investigating_2019; @williams_subtle_2016]. See Figure S\@ref(fig:heatmapsex).

Despite sex differences, there is also likely to be substantial overlap in many of the qualities and functions of female and male leaders across human societies. Leader sex was not a predictor of the majority of the measures investigated here (adjusting for other factors; see Table S\@ref(tab:universalvars)). When sex was a predictor (Table S\@ref(tab:variablevars)), descriptions including both female and male leaders provided relatively more supporting evidence for many key measures of leader qualities and functions, including group organization, resource distribution, ritual functions, and being brave and experienced and accomplished (see Figure S\@ref(fig:heatmapsex)).

@@

Leadership and social influence are expected to be associated with particular suites of individual qualities and functions, which potentially covary by cultural or social context, personality, or other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @bass_handbook_1990]. Our systematic cross-cultural results support many aspects of influential theories. Much of the evolutionary theoretical literature on leadership, however, has suggested social influence is most often achieved via one of two distinct strategies: dominance or prestige [for review see @garfield_review_2019; @garfield_evolutionary_2019]. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and social hierarchy and inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_j.k_dominance_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on distinctions between dominance and prestige suggests, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions, beyond dominance and prestige. Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, as well as their social networks, in order to organize group members, implement strategic actions, and provide prosocial services to the group, all while conforming to cultural norms. Dichotomizing this complexity into two distinct social strategies, although theoretically and empirically convenient, likely failes to capture much of the ethnographic reality of human leadership dynamics.

<!-- DISCUSSES PROSOCIALITY FEATURE -->
The prosociality feature of leader qualities, including generosity, interpersonal and oratory skills, and charisma as well as the role of cultural values shaping leadership styles, including proper behavior and having a favorable personality, generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @howell_role_2005]. The importance of prosociality  also corresponds to the @henrich_big_2015 Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

<!-- DISCUSSES SUCCESS FEATURE -->
Leaders, probably as a consequence of their ability to provide social benefits and impose social costs, are also characterized by their material, social, and reproductive success (bottom cluster Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature focused on associated outcomes of leadership and social hierarchy [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Several of these measures of success often co-occur. Among the Trobriands, horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

@@

@@

>>>>

Leaders were widely reported to receive material benefits, social status, and mating benefits but loss of material resources and social status were also widely reported, suggesting leadership to be a high-risk, high-return strategy (see Figure \@ref(fig:varsupportcb)). The results were similar for followers: followers benefited from increased material and social resources, but also often lost material and social resources. Followers were not reported to receive mating benefits, relative to leaders or other types benefits. Our results cannot determine if the benefits to leaders and followers outweighed their costs, but they do shed light on the social and biological pressures associated with leadership and social hierarchy. For both leaders and followers, individuals stood to lose what they were able to gain. See Figure \@ref(fig:varsupportcb).

All evolutionary models of leadership propose that the fitness benefits of leading outweigh the costs [e.g., @van_vugt_leadership_2008; @kantner_identifying_2010]. The relative abundance of evidence for leader and follower benefits vs. costs ($OR=`r signif(leader_bc_ratio$benefit_cost_OR, 2)`$ and $OR=`r signif(follower_bc_ratio$benefit_cost_OR, 2)`$, resp.), while not providing evidence of net fitness benefits, is nevertheless encouraging, especially since leaders and followers both had more evidence of mating and territory benefits than costs, which are key contributors to fitness, and leaders also had more evidence for inclusive fitness, material and social benefits, relative to costs. See Figure \@ref(fig:benefitcostratio).

@@

Relationships between leadership and high status, including causality and dependence, are puzzling and currently the focus of much work in evolutionary leadership studies [e.g., @van_vugt_dual_2019; @smith_leadership_2020; @cheng_why_2020; @cheng_dominance_2019; @von_rueden_dynamics_2019; @garfield_evolutionary_2019]. Drawing on a comparative-evolutionary perspective, @van_vugt_dual_2019 emphasize that although leadership and high social status do often covary, leaders are not always high-status individuals, exemplified by low-ranking individuals in nonhuman species that lead group movement or younger and lower status men and women who initiate collective actions. @cheng_dominance_2019 and @cheng_why_2020 emphasize, however, that among humans high social status may most often be a necessary but insufficient condition for achieving and maintaining positions of leadership. The "low status" young adults who initiate a collective action might be high status among their peers, within what @garfield_review_2019 [p. 71] described as the "intragenerational social hierarchy."

In our view, social context is critical to these discussions [see @wiessner_power_2010]. Leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social "statuses." In many cultures, for example, communities form work-groups and appoint "chiefs-for-a-day" who organize the work but have little status or authority otherwise [e.g., @macfarlan_competitive_2012]. @garfield_evolutionary_2019 found that leader organization variables clustered separately from the leader prestige variables, suggesting they are distinct. @garfield_investigating_2019 found that some leaders in recently settled hunter-gatherers had organizing responsibilities but unexceptional levels of respect. Some older individuals, on the other hand, were highly respected but had little social influence or authority.

In our results, high status was the most common quality of leaders at both the text record  (`r statusPercTxt`%) and culture level (`r statusPercCult`%) (Figure \@ref(fig:varsupport)); it was also the second most common benefit of leadership, and the loss of status was the most frequent cost (Figure \@ref(fig:varsupportcb)).
<!-- mentioned in this sample of the ethnographic record which includes discussions of chiefs, work group leaders, shamans, and age-sets, from a culturally diverse sample including foraging groups and kingdoms.  -->
This lends support to the notion of high social status often being a necessary condition of leader emergence [@cheng_why_2020]. Nevertheless, over two thirds of our text records on leadership did not mention status, and it was not one of our candidate universal dimensions of leadership. Interestingly, evidence for high status leaders varied by continental region, with more evidence for them from South American cultures (see Figure S\@ref(fig:heatmapregion)).

The dimensions of leadership predictive of high status leaders (see Figure \@ref(fig:highstatusplot)B) suggest the phenotype of such leaders resembles what many anthropologists refer to as "Big men" [@garfield_review_2019]: wealthy, accomplished older men who favor members of their kin group, and are highly regarded for their decision-making abilities, knowledge, and intelligence. High status leaders are able to direct benefits to their close relatives, likely due, in part, to their greater material and social wealth, which implicates leadership dynamics as a driver of the evolution and emergence of inequality [c.f., @sanday_female_1981; @shenk_status_2016; @price_social_2010; @borgerhoff_mulder_pastoralism_2010; @mattison_evolution_2016; @bowles_emergence_2010]. Our results did not strongly implicate high status leaders with particular functions, with the exception that punishment was modestly negatively associated with evidence for high status.

@smith_leadership_2020 speculate that, "leadership may...be more strongly correlated with high status in large, complex organizations, such as in corporations and governments (p. 2)." Evidence for high status leaders, however, did not substantially vary by group structure, which included levels of social organization ranging from sub-residential to supracommunity (see Figure S\@ref(fig:groupstatus)). This null result is evidence against the @smith_leadership_2020 hypothesis.

Overall, our results suggest that although high status leaders are common across cultures and all levels of social organization, only some leaders are high status, namely, those with a limited set of qualities such as wealth, experience, decision-making abilities, intelligence, and inter-personal skills, as well as older age and conformity to social norms.

@@

Only `r sum(all_data$demo_sex == 'female')` of the `r nrow(all_data)` text records discussed female leaders exclusively, and another `r sum(all_data$demo_sex == 'both')` explicitly discussed both female and male leaders. Leader sex was only associated with variation in seven of our `r length(all_study_vars)` dimensions of leadership (see Figure S\@ref(fig:heatmapsex)), and one feature variable (see Figure S\@ref(fig:heatmapfeaturesex)), indicating that most dimensions of leadership probably characterize leaders of both sexes.

When ethnographic texts described female leaders exclusively, they were relatively more likely to be described as married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits (Figure S\@ref(fig:heatmapsex)). Female leaders also had higher levels of evidence for the social, reproductive, and material success "feature" variable (Figure S\@ref(fig:heatmapfeaturesex)). Evidence for female leaders was also more likely to be found in kin groups and residential subgroups. This suite of characteristics is consistent with discussions of high status first-wives who gain and maintain increased social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985; @yanca_female_2004]. @garfield_investigating_2019 similarly found that there was positive assortative mating of high status leaders among elected leaders in a small-scale society with relative gender-egalitarianism. Overall, our resuls are also consistent with theoretical arguments emphasizing women as leaders of families [e.g., @garfield_evolutionary_2019; @hagen_leadership_2019].

@@

Leadership and social influence are expected to be associated with particular suites of individual qualities and functions, which potentially covary by cultural or social context, personality, or other inter-individual variation [@cheng_two_2013; @von_rueden_roots_2014; @hogan_personality_2020; @lukaszewski_extraversion_2015; @henrich_evolution_2001; @bass_handbook_1990].

Our systematic cross-cultural results support many aspects of influential evolutionary theories of leadership.

**Dominance and prestige**. Much of the evolutionary theoretical literature on leadership has suggested social influence is most often achieved via one of two distinct strategies: dominance or prestige [for review see @garfield_review_2019; @garfield_evolutionary_2019]. The dominance-prestige theory as formulated by @henrich_evolution_2001 significantly advanced evolutionary approaches to leadership and social hierarchy and inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2019]. @van_vugt_dual_2019 [p. 1], for example, "argue that there are two leadership styles: prestige style and dominance style," and @cheng_two_2013 [p. 119] in focusing on distinctions between dominance and prestige suggests, "human social stratification is characterized by the co-occurrence of both strategies." Exploratory cluster analyses reported here, however, suggest multiple features of leader qualities and leader functions, beyond dominance and prestige. Leaders likely rely a range of individual capacities, including physical, supernatural, and cognitive endowments, as well as their social networks, in order to organize group members, implement strategic actions, and provide prosocial services to the group, all while conforming to cultural norms. Dichotomizing this complexity into two distinct social strategies, although theoretically and empirically convenient, likely fails to capture much of the ethnographic reality of human leadership dynamics.

<!-- DISCUSSES PROSOCIALITY FEATURE -->
**Prosociality**. The prosociality feature of leader qualities, including generosity, interpersonal and oratory skills, and charisma as well as the role of cultural values shaping leadership styles, including proper behavior and having a favorable personality, generally correspond to a charismatic style of leadership that has been identified in industrialized settings and widely discussed in sociology and more recently emphasized in evolutionary psychology models of leadership [@den_hartog_culture_1999; @weber_distribution_1978; @grabo_charismatic_2016; @howell_role_2005]. The importance of prosociality  also corresponds to the @henrich_big_2015 Big-man mechanism, which suggests positive assortment of followers to prosocial leaders can facilitate the evolution of cooperative social norms and prestige-based leadership.
<!-- A few variables did not seem to fit prosocial theme, however, such as *Physical health*, *Physical formidable*, and *Drug consumption* (although drug use is often an important social activity). -->

<!-- DISCUSSES SUCCESS FEATURE -->
**Leader success**. Leaders, probably as a consequence of their ability to provide social benefits and impose social costs, are also characterized by their material, social, and reproductive success (bottom cluster Figure \@ref(fig:cluster)A), which is consistent with much of the theoretical literature focused on associated outcomes of leadership and social hierarchy [e.g., @von_rueden_mens_2016; @bowles_emergence_2010; @glowacki_role_2013]. Several of these measures of success often co-occur. Among the Trobriands, horticulturalists in Papua New Guinea, @ol06-037 [p. 47] describes relationships between polygynous marriages, increased wealth, and widespread social contacts:

@@

### Some notes

[NEED TO INTEGRATE]
These results provide a framework for further work on evolutionarily relevent costs and benfits of leadership, and suggests leaders-follower dynamics should continue to be prioritized in evolutionary models of demography, group dynamics, social complexity and inequality, cooperation and punishment, and resource exchange [c.f., @bird_variability_2019; @ember_our_2018; @micheletti_modelling_2020; @smith_cultural_2020; @birch2017philosophy; @marlowe_more_2008; @henrich_markets_2010; @mace_co-evolution_1998; @mace_phylogenetic_2005; @mattison_evolution_2016]

@@

Leadership dimensions for which leader sex significantly improved fit. Note that very few text records (N=`r sum(all_data$demo_sex=='both')`) explicitly mentioned both male and female leaders. Hence, although not depicted here, the confidence intervals for the estimated marginal means for this level are wide.
